---
title: The Effect of Dose and Delivery Method of Vitamin C on the Growth of Odontoblasts
  of the Incisor Tooth of the Guinea Pig
author: "John James"
date: "March 28, 2016"
output:
  word_document: default
  html_document:
    theme: journal
---

```{r setup, include=FALSE, echo=TRUE}
library(knitr)
library(kfigr)
library(printr)
read_chunk("toothgrowth.R")
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, fig.path = "../figures/", echo=TRUE, warning=FALSE)
```

```{r environment, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<environment>>
```
#Introduction
**Objectives:** The purpose of the study was to ascertain the degree to which delivery method and/or dosage of vitamin C affect growth of odontoblast cells in guinea pigs.

**Data Analysis:** Distributions of data were graphically summarized and two sample t-tests were conducted by dose, delivery method, and treatment to ascertain statistical significance of observations.

**Conclusions:** There is a positive correlation between vitamin C dose and incisor odontoblast cell growth in guinea pigs, regardless of delivery method.  Further, greater cell growth is associated with the administration of vitamin C with orange juice, at the lower doses (0.5mg and 1mg). There is no statistically significant difference in cell growth between delivery methods at the 2mg dose.

#Data
The data frame consisted of 60 observations of 3 variables.

1. Length - The length of the odontoblast cell in microns[3]
2. Supplement - The delivery method, either orange juice or an aqueous solution of ascorbic acid
3. Dose - Dose in milligrams per day. Values were 0.5, 1.0, and 2.0.

To facilitate data analysis, the data frame was copied and the dose was changed to a factor variable and the dose and supp variables were combined to create a third combined variable, hereinafter referred to as "treatment".
```{r copyDataFrame, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<copyDataframe>>
```

```{r descStats, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<descStats>>
```

```{r power, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<power>>
```

As shown in `r figr("tgStats", TRUE, type="Table")`, cell lengths range from `r tgStats$Min` to `r tgStats$Max` microns and have mean μ = `r round(tgStats$Mean,2)` and variance σ^2^ = `r round(tgStats$Var,2)`. There were `r tgStats$NAs` NA values observed in the data.

##Odontoblast Cell Growth and Dose, Delivery Method and Treatment
`r figr("boxPlots", TRUE, type="Figure")` (left) presents a positive correlation between dose and mean odontoblast cell growth with mean cell lengths of `r round(doseStats[[1,6]],2)`, `r round(doseStats[[2,6]],2)`, and `r round(doseStats[[3,6]],2)` microns for doses 0.5mg, 1mg, and 2mg respectively. `r figr("boxPlots", TRUE, type="Figure")` (right) shows greater cell growth across all doses for vitamin C administered with orange juice. Mean cell lengths were `r round(suppStats[[1,6]],2)` and `r round(suppStats[[2,6]],2)` microns for orange juice and the aqueous solution, respectively. Greater cell growth was also associated with the administration of orange juice at  the lower doses; however, there was no significant difference in cell growth between the delivery methods at the 2mg dose (`r figr("boxPlots", TRUE, type="Figure")` bottom).

##Data Distribution
`r figr("distributions", TRUE, type = "Figure")` evinces an approximately normal, albiet slightly left skewed, distribution of the observations for all doses and, with the exception of a spike in measurements at about 9 microns, for treatments where vitamin C was administered with orange juice. Measurements for the 0.5 and 2mg doses, as well as for those in which vitamin C was administered with the aqueous solution, are slightly right skewed normal.  The distribution of measurements for the 1mg dose are bimodal and nearly symmetrical.

#Methods
Two sample, unpaired, t-tests were conducted on a univariate basis to confirm the statistically significant differences in mean cell growth among the various doses and delivery methods. Having established results on a univariate basis, t-tests were conducted to ascertain any statistically significant differences in cell growth among the various treatments.

The following outlines the **assumptions** undertaken for this analysis:

1. Insofar as parametric t-tests are robust to assumptions of normality for small sample sets[4], normality was assumed for the total data set and for each subset.
2. A power π = 0.8 is assumed.  As such, differences in mean cell growth of `r round(del10,2)`, `r round(del20,2)`, and `r round(del30,2)` should be detectible at α = 0.5 and at assumed σ = `r round(tgStats[1,8])`, for sample sizes of 10, 20, and 30 respectively (`r figr("powerCurve", TRUE, type = "Figure")`).
3. Each pig was administered one, and only one combination of dose and delivery method, or treatment. As such, the t-tests conducted assumed unpaired samples.
4. Given the difference in variability among the various doses and delivery methods, unequal variances were assumed.

#Analysis
```{r ttest, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<t.tests>>
<<storeResults>>  
```
##Impact of Vitamin C Dose on Odontoblast Cell Growth
To ascertain the statistical difference in mean cell growth among the **doses**, let:

* u1 = mean cell growth for 0.5mg dose
* u2 = mean cell growth for 1mg dose
* u3 = mean cell growth for 2mg dose

`r figr("doseTest", TRUE, type="Table")` reveals differences in mean cell growth within confidence intervals that do not contain zero. Further, the p-values are significantly less than α = 0.5. Therefore, both null hypotheses are rejected.  As such, there is a statistically significant positive correlation between cell growth and dose. 

##Impact of Vitamin C Delivery Method on Odontoblast Cell Growth
To ascertain the statistical difference in mean cell growth among the **delivery methods**, let:

* u1 = mean cell growth for orange juice supplement
* u2 = mean cell growth for aqueous solution supplement

According to this test (`r figr("suppTest", TRUE, type="Table")`), the administration of vitamin C with orange juice produced a mean cell length `r suppTest$Delta` microns greater than that produced with the administration of vitamin C with the aqueous solution.  This is within the confidence interval that does not contain zero. Further, the p-value is significantly less than α = 0.5. Therefore, the null hypothesis is rejected.

##Impact of Treatment (Dose/Delivery) on Odontoblast Cell Growth
The following treatments were tested:

* Treatment 1: 0.5mg dose of Vitamin C with orange juice, where μ1 = mean cell length
* Treatment 2: 0.5mg dose of Vitamin C with aqueous solution, where μ2 = mean cell length 
* Treatment 3: 1mg dose of Vitamin C with orange juice, where μ3 = mean cell length
* Treatment 4: 1mg dose of Vitamin C with aqueous solution, where μ4 = mean cell length
* Treatment 5: 2mg dose of Vitamin C with orange juice, where μ5 = mean cell length
* Treatment 6: 2mg dose of Vitamin C with aqueous solution, where μ6 = mean cell length

Tests 1 and 2 (`r figr("treatmentTest", TRUE, type="Table")`)  show statistically significant differences in mean cell growth between the delivery methods at the lower doses (0.5mg and 1mg).  Test 3 indicates no statistically significant difference in mean cell growth between the delivery methods at the 2mg dose. 

#Conclusions
The analysis shows statistically different mean odontoblast cell growth at each **dose** level.  Higher doses directly correlate with greater cell growth, regardless of delivery method.

The impact of **delivery method** on odontoblast cell growth depends upon the dose. At the aggregate level, vitamin C administered with orange juice produces the greatest cell growth. The analysis shows that this is also true for the 0.5mg and 1mg dosing regimes.  On the other hand, there is no statistically significant difference in cell growth between the delivery methods at the higher (2.0mg) dose.  

#References
1. Crampton, E. W. (1947). The growth of the odontoblast of the incisor teeth as a criterion of vitamin C intake of the guinea pig. The Journal of Nutrition, 33(5): 491–504.
2. Maechler, M. (n.d.). R Manual. Retrieved from R Manual: https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html
3. Kuns, E. (2015, April 17). Bug 15953 - ToothGrowth {datasets} description improvement. Retrieved from Bugzilla: https://bugs.r-project.org/bugzilla3/show_bug.cgi?id=15953
4. Winter, J. d. (2013). Using the Student’s t-test with extremely small sample sizes. Practical Assessment, Research & Evaluation.


#Appendix

##Figures
```{r boxPlots, echo = FALSE, fig.height = 4, fig.width = 10, anchor = "Figure", message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<boxPlots>>
```

`r figr("boxPlots", TRUE, type="Figure")`: Odontoblast Cell Growth by Dose, Delivery & Treatment


```{r distributions, echo = FALSE, fig.height = 8, fig.width = 10, anchor = "Figure", message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<distributions>>
```

`r figr("distributions", TRUE, type="Figure")`: Distributions of Odontoblasts Cell Growth

```{r powerCurve, echo = FALSE, fig.height = 4, fig.width = 10, anchor = "Figure", message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<plotPowerCurve>>
```

`r figr("powerCurve", TRUE, type="Figure")`: Power Curve


##Tables
`r figr("tgStats", TRUE, type="Table")`: Descriptive Statistics for Tooth Growth Data Set
```{r tgStats, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE), anchor = "Table"}
tgStats
```


`r figr("doseTest", TRUE, type="Table")`: Two Sample t-Tests of Cell Growth by Dose
```{r doseTest, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE), anchor = "Table"}
doseTest
```

`r figr("suppTest", TRUE, type="Table")`: Two Sample t-Test of Cell Growth by Delivery Method
```{r suppTest, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE), anchor = "Table"}
suppTest
```

`r figr("treatmentTest", TRUE, type="Table")`: Two Sample t-Test of Cell Growth by Treatment
```{r treatmentTest, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE), anchor = "Table"}
treatmentTest
```

##Code
**Conduct t-tests**
```{r echo = TRUE, size = "small", message = F, eval = F}
<<t.tests>>
```